---
title: "Mapping DrugBank Compounds with UniChem"
output:
  html_document:
    theme: cosmo
    highlight: pygments
---

For more info or as a citation, please see:

+ Himmelstein DS, Brueggeman LA, Baranzini SE (2015) **Repurposing drugs on a heterogeneous network**. *ThinkLab*. [doi:10.15363/thinklab.4](https://dx.doi.org/10.15363/thinklab.4)

Specifically, this notebook corresponds to a dicussion on [unifying compound vocabularies using UniChem](//thinklab.com/discussion/unifying-drug-vocabularies/40).

First, we used [UniChem to match DrugBank compounds](//nbviewer.ipython.org/url/git.dhimmel.com/drugbank/unichem-map.ipynb) to external resources. We used the UniChem connectivity search which allows fuzzy matching.


```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(DT)
library(reshape2)
```

```{r}
drugbank.df <- file.path('data', 'drugbank.tsv') %>%
  read.delim(stringsAsFactors=FALSE) %>%
  dplyr::filter(type == 'small molecule') %>%
  dplyr::filter(grepl('approved', groups))

count.df <- file.path('data', 'mapping-counts.tsv') %>%
  read.delim(stringsAsFactors=FALSE, check.names=FALSE)

count.df <- drugbank.df %>%
  dplyr::rename(drugbank_name = name) %>%
  dplyr::left_join(count.df)
```

## The external sources that [we want to map to](//thinklab.com/discussion/unifying-drug-vocabularies/40)

```{r}
sources <- c('chembl', 'drugbank', 'fdasrs', 'pubchem', 'lincs')
```

## Percent of approved small molecules in DrugBank matched to external resource

```{r}
count.df %>%
  dplyr::select(one_of(sources)) %>%
  dplyr::summarise_each(funs(mean(. > 0) * 100)) %>%
  knitr::kable()
```

A small number of DrugBank approved small molecules (`r sum(count.df$drugbank == 0)`) do not map to DrugBank. This appears to occur because these compounds do not contain structural information in the DrugBank database.

## The number of compounds matching each approved small molecules in DrugBank

```{r}
count.df %>%
  dplyr::select(drugbank_id, drugbank_name, one_of(sources)) %>%
  DT::datatable()
```

## The distribution of compounds mapped per approved small molecule in DrugBank

```{r, message=FALSE}
count.df %>%
  dplyr::select(one_of(sources)) %>%
  reshape2::melt(variable.name = 'source', value.name = 'count') %>%
  ggplot(aes(count)) + theme_bw() +
    geom_histogram(binwidth = 1, origin = -0.5, alpha = 0.6, col='black') +
    facet_wrap(~ source, scales='free_x')
```

## Read the mapping file

```{r}
mapping.df <- file.path('data', 'mapping.tsv.gz') %>%
  read.delim(stringsAsFactors=FALSE)
```

## Each approved small molecule in DrugBank and the DrugBank IDs it matched to

```{r}
mapping.df %>%
  dplyr::filter(source_name == 'drugbank') %>%
  dplyr::inner_join(drugbank.df %>% dplyr::select(drugbank_id, type, groups)) %>%
  dplyr::group_by(drugbank_id, drugbank_name) %>%
  dplyr::summarise(
    n_matches = n(),
    matches = paste(src_compound_id, collapse = '|')
    ) %>%
  dplyr::filter(n_matches > 1) %>%
  DT::datatable()
```
